---
title: "Exploring Modern Difference-in-Differences Techniques with A Monte Carlo Simulation"
author: "Gregory Vander Vinne"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: false
---

```{r setup, include=TRUE, message=FALSE, warning=FALSE}
#Clear memory
rm(list = ls(all=T))


#Load packages
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
  
  tidyverse,  # Grammar for data + ggplot2
  did,        # Callaway and Sant'Anna
  did2s,      # Gardner
  data.table, # data tables
  knitr       # Print pretty tables
  
)

#Set some output options
knitr::opts_chunk$set(warning = FALSE, message = FALSE, fig.width = 10, fig.height = 7)
```

## Introduction

In my day job, I do a fair bit of work in the realm of causal inference. Typically, this involves looking at whether a particular policy resulted in improved outcomes for individuals. There are many different ways of doing this. One of the most common methods, or groups of methods, is difference-in-differences (DiD). DiD methods have been advancing rapidly in recent years following the discovery of a number of issues with two-way fixed effects (see Borusyak and Jaravel (2018), Sun and Abraham (2021), de Chaisemartin and D'Haultfoeuille (2020), and Goodman-Bacon (2021)).

Recently I have been exploring how different variations of difference-in-differences estimators that allow for staggered adoption behave under different circumstances. The {did2s} and {did} packages in R make this relatively easy to do with Monte Carlo simulations. While I have explored other estimators as well, in this post I wish to focus on two in particular. These two are the estimators proposed in Gardner (2021), and Callaway and Sant'Anna (2021). I will not be providing any proofs or getting into any of the math behind how these to methods work. Rather, I will be illustrating with a Monte Carlo simulations, how these different estimators behave (or misbehave) given different types of treatment effects and covariates. 

## A Hypothetical Scenario

I find simulation-based explanations easier to follow when they are explained in terms of a hypothetical real world scenario and we give the variables realistic values and names. For this simulation, let us imagine that a large company with many locations wants to measure whether adoption of a particular technology improves sales. Some of this company's locations have adopted the technology, while others have not. Furthermore, the locations that have adopted the technology have adopted it at different times and have chosen to adopt it of their own volition, which means that the treatment is not randomly assigned. Let us explore how we might measure the effects of the technology on sales using the methods of Callaway and Sant'Anna (2021) hereafter referred to as CS, Gardner (2021), hereafter referred to as Gardner, and TWFE.

### Homogeneous Treatment and Covariates with Fixed Effects

``` {r set parameters}
#Set number of units and periods
n.units <- 5000
n.periods <- 5

```

To begin, we have a balanced panel consisting of `r n.periods` periods (years) and `r n.units` units (store locations). Approximately 50%  of stores have not adopted the technology (i.e., they are untreated), while they other 50% have adopted the technology (i.e., they are treated at some point). Among these stores, approximately 1/4 adopted the technology in each of the last four years of our panel. In the first year, none of the stores have adopted the technology yet. For now, we will say that stores are treated in the first year that they adopt the technology and remain treated in all subsequent years. Additionally, there is one covariate, which is a dummy variable that indicates whether a store is located in an urban area or not. Approximately half of stores are in urban areas. Stores located in urban areas are twice as likely to adopt the technology as stores in non-urban areas; approximately 2/3 stores in urban areas adopt the technology, while 1/3 of stores in non-urban areas do. 

Lastly, we have revenue, which is determined as follows. For each store, revenue in each year is drawn from a random normal distribution with a mean of 1 000 000 and a standard deviation of 100 000, plus the effects of being located in an urban area and adopting the technology. The ceteris paribus effect of being located in an urban area is an additional 50 000 per year of revenue, while the ceteris paribus effect of the treatment is an additional 100 000 per year of income. This effect is completely homogeneous. 

```{r Generate Data, echo=TRUE}

#First column
myData <- as.data.frame(rep(1:n.units, n.periods))
colnames(myData) <- "id"

#For reproducibility
set.seed(97)
#Add other columns
myData <- myData %>%
  arrange(by=id) %>% 
  cbind(year = rep(2019:(2019+n.periods-1),n.units)) %>% #Add periods
  group_by(id) %>%
  mutate(g = case_when(
    id %in% (1:n.units/2) ~ 0, #50% untreated
    id %in% ((n.units/2+1):(n.units/2+n.units/(n.periods-1)/2)) ~ 2020,
    id %in% ((n.units/2+n.units/(n.periods-1)/2+1):(n.units/2+n.units*2/(n.periods-1)/2)) ~ 2021,
    id %in% ((n.units/2+n.units*2/(n.periods-1)/2+1):(n.units/2+n.units*3/(n.periods-1)/2)) ~ 2022,
    TRUE ~ 2023
  ), # Assign to treatment groups
         urban = case_when(g==0 ~ sample(c(0,0,1),1),
                           TRUE ~ sample(c(0,1,1),1)) # Urban locations twice as likely to be treated
  ) %>%
  rowwise() %>%
  mutate(treated = case_when(g<=year & g!=0 ~ 1,
                             TRUE ~ 0),
         revenue = (rnorm(1, mean = 1000000, sd =100000) + #Average revenue of 1M plus...
           50000*urban + # additional 100k if in urban area
           100000*treated # treatment effect: 100k 
           ) 
         ) 
```

Now, to ensure that the reader has a good comprehension of the simulated data set, I provide a few tables to provide some insight into the data. The first table shows what the data set looks like for one treated and one untreated store. 

```{r A Glimpse of the Data, echo=TRUE}
#Filter down to the five observations for one treated individual
myData %>% 
  ungroup() %>%
  filter(g==0, urban == 0)%>%
  slice_head(n=5) %>%
  #Stack on top of five obs for one untreated individual
  rbind(
    #Filter down to the five observations for one individual treated in 2017
    myData %>%
      ungroup() %>%
      filter(g==2021, urban==1)%>%
      slice_head(n=5)
  ) %>%
  mutate(revenue = scales::dollar(revenue)) %>%
  #Print as a table
  kable(
    align = "l",
    caption = "Data for One Store That Adopts the Technology and One that Does Not",
    col.names = c("ID","Year","Treatment Group","Urban Dummy","Treated Yet", "Revenue")
  )
```

The table below shows mean revenue by treatment status and whether a store is located in an urban area. 

``` {r Revenue by Treatment and Urban, echo = TRUE}
myData %>%
  group_by(treated, urban) %>%
  summarise(Revenue = mean(revenue) %>% scales::dollar()) %>%
  kable(
    align = "l",
    caption = "Mean Revenue by Treatment Status and Whether Stores are Urban",
    col.names = c("Treated","Urban Dummy", "Mean Revenue")
  )


```

Finally, this table shows the number of units with each treatment group / urban dummy combination.


``` {r Count by Treatment Group and Urban, echo = TRUE}
myData %>%
  group_by(g, urban) %>%
  summarise(Count = n()) %>%
  kable(
    align = "l",
    caption = "In Each Treatment Group by Urban Dummy ",
    col.names = c("Treatment Group", "Urban Dummy", "Count")
  )

```

Now that the housekeeping is out of the way let's look at how the different estimators perform in this scenario. The {did2s} package makes it very easy for us to compare the results. As can be seen from the results below, the three methods all provide similar, unbiased estimates of the effect of the treatment. Note that we have not explicitly controlled for or conditioned on the the urban covaraite in within any of the three methods this. In the cases of TWFE and Gardner (2021), the unit-level fixed effects will deal with the covariate in this case, and in CS, the 'differencing' will deal with it. This is because although the covariate effects the **level** of the outcome variable, it does not effect the **dynamics** of the outcome variable - it has the same value and same effect in all periods.    

``` {r Estimates For Scenario 1, echo = TRUE}
#Without conditioning on covariates
output <- event_study(
  yname = "revenue",  
  tname = "year",       
  idname = "id",        
  gname = "g",      
  data = myData
) %>%
  filter(estimator %in% c("TWFE", "Gardner (2021)", "Callaway and Sant'Anna (2020)"))

#Plot the results
plot_event_study(output) 


```

### Heterogeneous Treatment Effects

``` {r Effect Parameters}
#Effect that treatment has in each year
e_20 <- 50000
e_21 <- 100000
e_22 <- 150000
e_23 <- 200000


```

Now, let us consider a case in which the effects of the treatment are heterogeneous. That is, the effects are not the same for all groups and periods. We will make it so that the treatment effect is +`r e_20` in revenue in 2020, `r e_21` 2021, `r e_22` in 2022 and `r e_23` in 2023. Perhaps this steady increase in the benefit from the technology arises from some sort of network  effect -  as more people/firms/locations adopt the technology, it becomes more effective. Note that it does not become more effect because units have been in the 'treated' state for more years. It becomes more effective as more calendar years pass, regardless of how long a given unit has been treated.

``` {r Estimates For Scenario 2, echo = TRUE}

true_effect <- data.table(
  term = -4:3,
  true_effect = c(rep(0,4),
               (e_20+e_21+e_22+e_23)/4,
               (e_21+e_22+e_23)/3,
               (e_22+e_23)/2,
               e_23)
)


myData2 <- myData %>% 
  mutate(revenue = rnorm(1, mean = 1000000, sd =100000) +
           50000*urban +
           #treatment effect dependent on year
           case_when(
             year == 2019 ~ 0,
             year == 2020 ~ treated*e_20,
             year == 2021 ~ treated*e_21,
             year == 2022 ~ treated*e_22,
             year == 2023 ~ treated*e_23
           )
         ) 

#Without conditioning on covariates
output <- event_study(
  yname = "revenue",  
  tname = "year",       
  idname = "id",        
  gname = "g",      
  data = myData2
) %>%
  filter(estimator %in% c("TWFE", "Gardner (2021)", "Callaway and Sant'Anna (2020)"))

output <- left_join(output, true_effect, by = "term")

#Plot the results
plot_event_study(output) + 
  geom_line(aes(x = term, y=true_effect), color = "red")



```

#### Economic Shock?


### Covariates with Dynamic Effects

Let us return to a situation in which the effects of treatment are not heterogeneous. As in the first simulation, we will say that adopting the technology has the effect of increasing revenues by 100,000 per year in all years. However, we will introduce a covariate that does not have the same effect in all periods. Let's imagine that this is a Canadian company with approximately half of its locations in Canada and half in the USA. Initially, in 2019, there is a benefit to being located in Canada, perhaps because of better brand awareness. In 2019, being located in Canada has the effect of boosting revenues by 50 000, holding all else equal. However, as time goes on, the brand grows in popularity in the USA, perhaps competitors in the USA start closing down, or the brand just becomes more of a fad in the USA than in Canada. As a result, in 2020, beinging located in the united states has no benefit, while in 2021 onwards, being located in the United States has the effect of **increasing** revenues by 50 000. Furthermore, Canadian locations are less likely to adopt the technology than American locations, meaning that treatment is correlated with this variable. We will say that American locations are three times as likely to adopt the treatment as Canadian locations. 

Therefore, although the value of this covariate is constant over time, its effect is not constant. Rather than merely affecting the **level** of the outcome variable, this covariate affects the **dynamics** of the outcome variable. First, let us compute the three different estimators that we have been discussing, without explicitly controlling for or conditioning on this new covariate in any of them.  

``` {r Estimates For Scenario 2, echo = TRUE}
myData <- myData %>%
  mutate()




```




### Covariates with Changing Values


### Treatment that 'Switches Off'


